Designing a high-performance boundary element library with OpenCL and Numba

### 2021-04-13 main.tex
working through opencl section
Started performance section
editing up to simd section

\input{intro}
\input{bem}
\input{bempp_overview}
\input{opencl}

@@

\bibitem{Bempp-cl}
T. Betcke and M. W. Scroggs, ``Bempp-cl: A fast Python based just-in-time compiling boundary element library,''
{\it Journal of Open Source Software}, vol. 6, no. 59, p. 2879, 2021.
\textsc{doi}: \href{https://dx.doi.org/10.21105/joss.02879}{10.21105/joss.02879}
\end{thebibliography}

>>>>

\input{intro}
\input{bem}
\input{bempp_overview}
\input{opencl}
\input{numba}

@@

\bibitem{Bempp-cl}
T. Betcke and M. W. Scroggs,
``Bempp-cl: A fast Python based just-in-time compiling boundary element library,''
{\it Journal of Open Source Software}, vol. 6, no. 59, p. 2879, 2021.
\textsc{doi}: \href{https://dx.doi.org/10.21105/joss.02879}{10.21105/joss.02879}.

\bibitem{pyopencl}
A. Kloeckner, N. Pinto, Y. Lee, B. Catanzaro, P. Ivanov, and A. Fasih,
``PyCUDA and PyOpenCL: A scripting-based approach to GPU run-time code generation,''
{\it Parallel Computing}, vol. 38, no. 3, pp. 157--174, 2012.
\textsc{doi}: \href{https://dx.doi.org/10.1016/j.parco.2011.09.001}{10.1016/j.parco.2011.09.001}.

\bibitem{opencl}
OpenCL, \url{https://www.khronos.org/opencl/}.
\end{thebibliography}

### 2021-04-13 opencl.tex
editing up to simd section

OpenCL (\url{https://www.khronos.org/opencl/}) is a heterogeneous compute standard for CPUs, GPUs, FPGAs, and other types of devices that provide conformant drivers. At its core, OpenCL executes compute kernels that can be written in C99, or (more recently) in C++. The current version of OpenCL is 3.0, though the most widely implemented standard is OpenCL 1.2, which Bempp-cl uses.

@@

\begin{itemize}
	\item Test and trial indices of triangles over which to be integrated.
	\item Signs of the normal directions for the spaces.
	\item Test and trial grid as flat floating point array, defining each triangle through nine floating point numbers, specifying the $(x, y, z)$ coordinates of each of the three nodes of a triangle.
	\item Test and trial connectivity information that stores the indices of each node for each triangle.
	\item Test and trial mappings of local triangle degrees of freedom to global degrees of freedom.
	\item Test and trial basis function multipliers for each triangle.
	\item Quadrature points and quadrature weights.
	\item A buffer that contains the global assembled matrix.
	\item Additional kernel parameters, such as the wavenumber for Helmholtz problems.
	\item Number of test and trial degrees of freedom.
	\item A single byte that is set to one if the test and trial grids are disjoint.
\end{itemize}
The actual kernel definition is shown in \cref{fig:kernel_definition}.
Before the kernel can be launched, it needs to be configured and just-in-time compiled. Kernel configuration happens through C-style preprocessor definitions that are passed through the just-in-time compiler. These include the names of the test and trial space, the name of the function that evaluates the integral kernel, choosing between single and double precision types, and for SIMD enhanced kernel the vector length of the SIMD types.
For example, in Figure \ref{fig:kernel_definition} all floating point types have the name \ocl{REALTYPE}. This is substituted with either \textit{float} or \textit{double} during just-in-time compilation.

Each work-item computes via numerical quadrature all interactions via the integral kernel of basis functions on the trial element with basis functions on the test element. Before summing the result into the global result buffer, the kernel checks via the connectivity information if the test and trial triangle are adjoint or identical. For this it simply checks if at at least one of the node indices of the test triangle is identical to one of the node indices of the trial triangle. If this is true and the grids are not disjoint, the result of the kernel is discarded and not summed back into the global result buffer. The reason is that for adjoint triangles a separate singular quadrature rule is used. The effect is that a few work-items do work that is discarded. However, in a grid with $N$ elements the number of triangle pairs requiring a singular quadrature rule is $\mathcal{O}(N)$, while the total number of interactions is $N^2$. Hence, only a tiny fraction of work-items are discarded.

>>>>

OpenCL \cite{opencl} is a heterogeneous compute standard for CPUs, GPUs, FPGAs, and other types of devices that provide conformant drivers. At its core, OpenCL executes compute kernels that can be written in C99, or (more recently) in C++. The current version of OpenCL is 3.0, though the most widely implemented standard is OpenCL 1.2, which Bempp-cl uses.

@@

\begin{itemize}
	\item Test and trial indices of triangles over which to be integrated.
	\item Signs of the normal directions for the spaces.
	\item Test and trial grid as flat floating point array, defining each triangle through nine floating point numbers, specifying the $(x, y, z)$ coordinates of each of the three nodes of a triangle.
	\item Test and trial connectivity information that stores the indices of each node for each triangle.
	\item Test and trial mappings of local triangle degrees of freedom to global degrees of freedom.
	\item Test and trial basis function multipliers for each triangle.
	\item Quadrature points and quadrature weights.
	\item A buffer that contains the global assembled matrix.
	\item Additional kernel parameters, such as the wavenumber for Helmholtz problems.
	\item Number of test and trial degrees of freedom.
	\item A single byte that is set to one if the test and trial grids are disjoint.
\end{itemize}

An example kernel definition is shown in \cref{fig:kernel_definition}.
Before the kernel can be launched, it needs to be configured and just-in-time compiled. Kernel configuration happens through C-style preprocessor definitions that are passed through the just-in-time compiler. These include the names of the test and trial space, the name of the function that evaluates the Green's function, whether we are using single or double precision types, and (for SIMD enhanced kernels) the vector length of the SIMD types.
For example, in \cref{fig:kernel_definition} all floating point types have the name \ocl{REALTYPE}. This is substituted with either \ocl{float} or \ocl{double} during just-in-time compilation.

Each work-item computes (using numerical quadrature) all interactions of basis functions on the trial element with basis functions on the test element. Before summing the result into the global result buffer, the kernel checks via the connectivity information if the test and trial triangles are adjacent or identical (see \cref{fig:triangles}). To do this, it simply checks if at at least one of the node indices of the test triangle is equal to one of the node indices of the trial triangle. If this is true and the grids are not disjoint, the result of the kernel is discarded and not summed back into the global result buffer: for these triangles, separate singular quadrature rules need to be used. The effect is that a few work-items do work that is discarded. However, in a grid with $N$ elements, the number of triangle pairs requiring a singular quadrature rule is $\mathcal{O}(N)$, while the total number of interactions is $N^2$. Hence, only a tiny fraction of work-items are discarded.
